Parsing Postfix log files

### 2008-05-20 doc/logparser-background.tex
Corrections to sections 1 and 2 from my proofreading.

This section provides background information helpful in understanding the
remainder of the paper.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
introduction to Postfix, concentrating on the topics most relevant to this
paper, namely Postfix anti-spam restrictions and policy servers.  The
assumptions made in designing and implementing the parser are explained, as
are the conventions used in this paper.  Other projects which attempt to
parse Postfix log files are summarised (full details are available in
appendix~\refwithpage{other-parsers}), finishing with a review of
previously published research in this area.

@@

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are beneficial in the fight against spam.  The
program parses Postfix log files and populates a database with the data
gleaned from those log files, providing a consistent and simple view of the
log files which future tools can utilise.  The gathered data can then be
used to optimise current anti-spam measures, provide a baseline to test new
anti-spam measures against, or to produce statistics showing how effective
those measures are.

A short example of the optimisation possible using data from the database
is determining which Postfix restrictions reject the highest number of
mails:

@@

Another example is determining which restrictions are not effective: this
shows which restrictions had fewer than 100 hits on the last log file
parsed.

@@

These database queries yield summary statistics about the efficiency of
spam avoidance techniques that is far less feasible to assess directly from
log files without prior pre-processing into a database in the fashion
proposed, implemented and tested herein.

@@

    \item The majority of programming languages have existing code
        available providing access to databases, so the majority of
        languages will be able to access the results obtained by the
        parser.  If this parser was written to be utilised as shared code
        there would be little or no difficulty in using it when writing in
        the programing language it is implemented in, but every other
        programing language wishing to utilise it would require an
        interface layer to be written.  This requirement would drastically
        reduce the viability of any project wishing to build upon the work
        of this project.

    \item Databases provide complex querying and sorting functionality to
        the user without requiring large amounts of programming.  All
        databases provide a program, of variable complexity and
        sophistication, which can be used for ad hoc queries with minimal
        investment of time.

@@

            \item If the database supports it, access can be granted on a
                fine-grained basis so that the finance department can
                produce invoices, the helpdesk can run limited queries as
                part of dealing with support calls, and the administrators
                can have full access to the data.

@@

The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and
viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
adhered to the principle of \textit{Be liberal in what you accept, and
conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
unfortunately that principle was written in a friendlier time.  Given the
deluge of spam that mail servers are subjected to daily, a more appropriate
maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
the strongest restrictions you can; relax the restrictions and adherence
only when legitimate mail is impeded.\/}  It's not as friendly, nor as
catchy, but it more accurately reflects the current situation.} A typical
\SMTP{} conversation resembles the following (the lines starting with a
three digit number are sent by the server, all other lines are sent by the
client):

@@

A detailed description of \SMTP{} is beyond the scope of this document:
introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.

@@

Postfix is a \MTA{} with the following design aims (in order of
importance): security, flexibility of configuration, scalability, and high
performance.  It features extensive, extensible, optional anti-spam
restrictions, allowing an administrator to deploy those restrictions which
they judge suitable for their site's needs, rather than a fixed set chosen
by Postfix's author.  These restrictions can be selectively applied,
combined and bypassed on a per-client, per-recipient or per-sender basis,
allowing varying levels of severity and/or permissiveness.  Postfix
leverages simple lookup tables to support arbitrarily complicated
user-defined sequences of restrictions and exceptions, with policy
servers\footnote{Policy servers will be explained in \sectionref{policy
servers}.} as the ultimate in flexibility.  Administrators can also supply
their own rejection messages to make it clear to senders why exactly their
mail was rejected.  Unfortunately this flexibility has a cost: complexity
in the log files generated.  While it is easy to use standard Unix text
processing utilities to determine the fate of an individual email,
following the journey an email takes through Postfix can be quite
difficult.  The log files tend to follow a 90\%-10\% pattern: 90\% of the
time the journey is simple, but the other 10\% of the time requires 90\% of
the code.\footnote{These numbers don't have a solid scientific basis,
they're based on gut feeling from writing and debugging the software.}

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
principle was also followed when designing the parser (\sectionref{parser
design}), notably having separate actions rather than one huge action.} and
is separated into various component programs to perform the tasks required
of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
full details can be found at~\cite{postfix-overview}.  Each log line
contains the name of the Postfix program which produced it, and this
information is used when determining which rules should be used to parse
each log line (see \sectionref{rule characteristics} for details).  This
design has positive security implications also: those components which
interact with other hosts are not privileged,\footnote{Privilege in this
case means the power to perform actions that are limited to the
administrator, and not available to ordinary users.} so bugs in those
components will not give an attacker extra privileges; those components
which are privileged do not interact with other hosts, making it much more
difficult for an attacker to exploit any bugs which may exist in those
components.

@@

Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
smtpd_per_user_control, policy-servers}, the following is a brief overview
only.

Postfix uses one restriction list (containing zero or more restrictions)
for each stage of the \SMTP{} conversation: client connection, HELO
command, MAIL FROM command, RCPT TO commands, DATA command, and end of
data.  The appropriate restriction list is evaluated for each stage
(evaluation will be explained shortly), though by default the restriction
lists for client connection, HELO and MAIL FROM commands will not be
evaluated until the first RCPT TO command is received, because some clients
don't deal properly with rejections before RCPT TO\@; a benefit of this
delay is that Postfix has more information available when logging the
rejection.  Postfix uses simple lookup tables as the deciding factor when
evaluating some restrictions, e.g.\newline
\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\newline The line above can be broken down as follows:

@@

The restriction \texttt{check\_client\_access} checks whether the \IP{}
address of the connected client is found in the file and returned the
associated action if found; the method of searching the file is dependant
on the type of the file (\texttt{cidr} in the example) --- see
\cite{postfix-lookup-tables} for more details.  Other restrictions
determine their result by consulting external sources, e.g.\
\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
address of the client is present in the \RBL{} \texttt{rbl.example.com},
rejecting the command if the client is listed.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible, for full details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue processing the remainder of the restriction list, allowing more
specific cases to be used as exceptions to more general cases.  When the
result is the name of another restriction Postfix will evaluate the new
restriction, allowing restrictions to be chosen based on the client \IP{}
address, HELO hostname, sender address, recipient address, etc.  The
administrator can define new restrictions as a list of existing
restrictions, allowing arbitrarily long and complex sequences of lookups,
restrictions and exceptions.  Postfix tries to protect the administrator in
as far as is reasonable, e.g.\ the restriction
\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
because the parameter it checks (the hostname given in the HELO command) is
under the control of the remote client.  Despite this, it is possible for
the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
all users --- the administrator must remember that \textit{With great
power comes great responsibility}.

This description is necessarily brief, for further details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.

@@

A policy server~\cite{policy-servers} is an external program that accepts
state information from Postfix for each \SMTP{} command not rejected by an
earlier restriction; the policy server can utilise that state information
to implement whatever logic is required.  For example some users can be
restricted to sending mail on the third Tuesday after pay day only --- this
example may actually be useful in a payroll system to prevent problems from
spam or (worse) phishing mails with faked sender addresses.  More commonly
encountered scenarios are:

@@

    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
        \SPF{}\label{spf introduction} records specify which mail servers
        are allowed to send mail claiming to be from a particular domain.
        The intention is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter} and
        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of \SMTP{}\@.

@@

    \item Using a scoring system such as
        Policyd-weight~\cite{policyd-weight} where tests accumulate points
        against the sending system --- if the eventual score is too high
        the mail is rejected.  Postfix's restrictions are binary, and the
        administrator must manually whitelist clients if they are to bypass
        a restriction; using a threshold which requires hitting several
        restrictions frees the administrator from whitelisting clients
        which fall foul of one restriction only.

@@

@@

\end{tabular}

@@

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that log files produced by Postfix are not fixed; they vary
widely from host to host, depending on the set of restrictions chosen by
the administrator.  With this in mind, one of the parser's design aims was
to make adding new rules as easy as possible, to enable administrators to
properly parse their log lines.  To enable this the parser is divided into
three parts:

@@

    \item [rules] Rules match individual log lines and determine which
        actions will be executed.  Rules provide an easily extensible
        method of associating log lines and actions.  Rules are described
        in detail in \sectionref{rules}.

    \item [actions] Actions invoked to deal with a log line once it has
        been identified by the rules: modifying data structures, handing
        complications, and saving data to the database.  The actions
        perform the work of reconstructing the journey a mail takes through
        Postfix.  Full details of actions can be found in
        \sectionref{parsing-algorithm}, particularly
        \sectionref{actions-in-detail} and \sectionref{adding new actions}.

    \item [framework] The framework is responsible for loading rules,
        managing data structures so actions can retrieve them, finding the
        rule which matches each log line, invoking the correct action, etc.
        The framework provides the structure which actions and rules plug
        into.  The framework is not described in detail in this document
        because it is not particularly interesting, but there is some
        coverage in \sectionref{framework}.

@@

Decoupling the parsing rules from the associated actions allows new rules
to be written and tested without requiring modifications to the parser
source code (significantly lowering the barrier to entry for new or casual
users who need to parse new log lines), and greatly simplifies both actions
and rules.  Decoupling also creates a clear separation of functionality:
rules handle low level details of identifying log lines and extracting data
from a log line, whereas the actions handle the higher level details of
following the path a mail takes through Postfix, assembling the required
data before storing it, dealing with complications arising, etc.

@@

@@

\subsection{Conventions used in the paper}

The words \textit{connection\/} and \textit{mail\/} are often used
interchangeably in this paper; in general the word used was chosen based on
the context it appears in.

@@

Ten other parsers have been reviewed in
appendix~\refwithpage{other-parsers} as part of the background research for
this project.  None of the reviewed parsers perform the type of advanced
parsing and log correlation described here; all are intended to perform a
specific parsing and reporting task, rather than be a generic parser,
extracting data and leaving generation of reports from the data to other
programs.  Some parsers save the data extracted to a data store but the
majority discard all data once they have finished running, making
historical analysis impossible.  The other parsers reviewed all produce a
report of greater or lesser complexity and detail, whereas the program
described here doesn't attempt to produce a report at all; that
responsibility is deferred to a separate program, to be developed later.
The parsing algorithm and program described here are designed to enable
much more detailed log analysis by providing a stable platform for
subsequent programs to develop upon.

@@

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
different from this parser: it attempts to present correlated log data in a
form suitable for a systems administrator to search using the myriad of
standard Unix text processing utilities already available, producing both a
\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
support both \CSV{} and Berkeley DB appears to have been a limiting factor:
\CSV{} is a very simple format where a record is stored in a single line,
with fields separated by a comma or other punctuation symbol.  Problems
with \CSV{} files include the difficulty in escaping separators, providing
multiple values for a field (e.g.\ multiple recipients), adding new fields,
and lack of a bundled schema describing the fields.

Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
\SQL{} databases which store arbitrary tuples; in the main table the key is
an integer referred to by secondary tables, and the value used is a \CSV{}
line containing all of the data for that row.  The secondary by-sender,
by-recipient, by-date and by-\IP{} tables use the
sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
referring to the main table is the value.  In effect this re-implements
\SQL{} foreign keys, but without the functionality offered by even the most
basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
requires custom code to search on some combination of the above, though the
authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
of the three output formats

The data stored is limited to time and date, hostname and \IP{} address of
client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
codes or remote servers\footnote{A single mail may be sent to multiple
remote servers if it was addressed to recipients in different domains, or
Postfix needs to try multiple servers for one or more recipients.} is not
explained in the paper; experimental observation shows that multiple
records are added when there are multiple recipients (sadly the records are
not associated in any way), and presumably the same approach is taken when
there are multiple destination servers.

The schema used with the MySQL database is undocumented, but at least it's
possible to discover the schema with an \SQL{} database, unlike with
Berkeley DB\@; all \SQL{} databases embed the database schema into the
database and provide commands for displaying it.  Berkeley DB does not
embed a schema, as there is neither requirement nor benefit; it only
provides \textbf{(key, value)} pairs, so any additional structuring of the
data is imposed by the application, thus it behoves the application to
document this structure.

@@

This section has provided background information on several topics relevant
to the remainder of the paper.  It started with the motivation behind the
project, continuing with explanations of:

@@

    \item A description of a new parser's novel three section design.

@@

    \item A brief comparison of this parser versus other Postfix parsers.

>>>>

This section provides background information helpful in understanding the
remainder of the document.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
introduction to Postfix, concentrating on the topics most relevant to this
document, namely Postfix anti-spam restrictions and policy servers.  The
assumptions made in designing and implementing the parser are explained, as
are the conventions used in this document.  Other projects which attempt to
parse Postfix log files are summarised (full details are available in
appendix~\refwithpage{other-parsers}), finishing with a review of
previously published research in this area.

@@

This document and the program it describes are part of a larger project to
optimise a mail server's Postfix restrictions, generate statistics and
graphs, and provide a platform on which new restrictions can be trialled
and evaluated to determine if they are beneficial in the fight against
spam.  The program parses Postfix log files and populates a database with
the data gleaned from those log files, providing a consistent and simple
view of the log files which future tools can utilise.  The gathered data
can then be used to optimise current anti-spam measures, provide a baseline
to test new anti-spam measures against, or to produce statistics showing
how effective those measures are.

A snippet of \SQL{} provides a short example of the optimisation possible
using data from the database is determining which Postfix restrictions
reject the highest number of mails:

@@

XXX GIVE SAMPLE OUTPUT FOR THE SNIPPET ABOVE\@.

Another example is determining which restrictions are not effective: this
shows which restrictions had fewer than 100 hits on the last log file
parsed, and the percentage of total rejections each restriction represents.

@@

These database queries yield summary statistics about the efficiency of
spam avoidance techniques; statistics that are far less feasible to assess
directly from log files without prior pre-processing into a database in the
fashion proposed, implemented and tested herein.

@@

    \item XXX REWRITE THIS PARAGRAPH\@.  The majority of programming
        languages have existing code available providing access to
        databases, so the majority of languages will be able to access the
        results obtained by the parser.  If this parser was written to be
        utilised as shared code there would be little or no difficulty in
        using it when writing in the programing language it is implemented
        in, but every other programing language wishing to utilise it would
        require an interface layer to be written.  This requirement would
        drastically reduce the viability of any project wishing to build
        upon the work of this project.

    \item Databases provide complex querying and sorting functionality to
        the user without requiring large amounts of programming.  All
        databases provide a program, of varying complexity and
        sophistication, which can be used for ad hoc queries with minimal
        investment of time.

@@

            \item If the database supports it, access can be granted on a
                fine-grained basis, e.g.\ allowing the finance department
                to produce invoices, the helpdesk to run limited queries as
                part of dealing with support calls, and the administrators
                to have full access to the data.

@@

The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity of the protocol many virus and/or spam sending programs fail
to implement it properly, so requiring strict adherence to the protocol
specification is beneficial in protecting against spam and
viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
adhered to the principle of \textit{Be liberal in what you accept, and
conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
unfortunately that principle was written in a friendlier time.  Given the
deluge of spam that mail servers are subjected to daily, a more appropriate
maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
the strongest restrictions you can; relax the restrictions and adherence
only when legitimate mail is impeded.\/}  It's not as friendly, nor as
catchy, but it more accurately reflects the current situation.} A typical
\SMTP{} conversation resembles the following (the lines starting with a
three digit number are sent by the server, all other lines are sent by the
client):

@@

A detailed description of \SMTP{} is beyond the scope of this document:
introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02},
the definitive references are~\cite{RFC821, RFC2821}.

@@

Postfix is a \MTA{} with the following design aims (in order of
importance): security, flexibility of configuration, scalability, and high
performance.  It features extensive, extensible, optional anti-spam
restrictions, allowing an administrator to deploy those restrictions which
they judge suitable for their site's needs, rather than a fixed set chosen
by Postfix's author.  These restrictions can be selectively applied,
combined and bypassed on a per-client, per-recipient or per-sender basis,
allowing varying levels of stricture and/or permissiveness.  Postfix
leverages simple lookup tables to support arbitrarily complicated
user-defined sequences of restrictions and exceptions, with policy
servers\footnote{Policy servers will be explained in \sectionref{policy
servers}.} as the ultimate in flexibility.  Administrators can also supply
their own rejection messages to make it clear to senders why exactly their
mail was rejected.  Unfortunately this flexibility has a cost: complexity
in the log files generated.  While it is easy to use standard Unix text
processing utilities to determine the fate of an individual email,
following the journey an email takes through Postfix can be quite
difficult.  The log files tend to follow a 90\%-10\% pattern: 90\% of the
time the journey is simple, but the other 10\% of the time accounts for
90\% of the code in the parser.\footnote{These numbers don't have a solid
scientific basis, they're based on gut feeling from writing and debugging
the software.}

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
principle was also followed when designing the parser (\sectionref{parser
design}), notably in having separate actions rather than one huge action.}
and is separated into various component programs to perform the tasks
required of an \MTA{}\@: receive mail, send mail, local delivery of mail,
etc. --- full details can be found at~\cite{postfix-overview}.  Each log
line contains the name of the Postfix component which produced it, and this
information is used when determining which rules should be used to parse
each log line (see \sectionref{rule characteristics} for details).  This
design has positive security implications also: those components which
interact with other hosts are not privileged,\footnote{Privilege in this
case means the power to perform actions that are limited to the
administrator, and not available to ordinary users.} so bugs in those
components will not give an attacker extra privileges; those components
which are privileged do not interact with other hosts, making it much more
difficult for an attacker to exploit any bugs which may exist in those
components.

@@

\label{Mixing and matching Postfix restrictions}

Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
smtpd_per_user_control, policy-servers}; the following is a brief overview
only.

Postfix uses one restriction list (containing zero or more restrictions)
for each stage of the \SMTP{} conversation: client connection, HELO
command, MAIL FROM command, RCPT TO commands, DATA command, and end of
data.  The appropriate restriction list is evaluated for each stage
(evaluation will be explained shortly), though by default the restriction
lists for client connection, HELO and MAIL FROM commands will not be
evaluated until the first RCPT TO command is received, because some clients
don't deal properly with rejections before RCPT TO\@; a benefit of this
delay is that Postfix has more information available when logging
rejections.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible as described
in~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.} The
meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue processing the remainder of the restriction list, allowing more
specific cases to be used as exceptions to more general cases.  When the
result is the name of another restriction Postfix will evaluate the new
restriction, allowing restrictions to be chosen based on the client \IP{}
address, HELO hostname, sender address, recipient address,
etc.\footnote{E.g.\ the administrator may require that all clients on the
local network have valid \DNS{} entries, to prevent people sending mail
from unknown machines.}  The administrator can define new restrictions as a
list of existing restrictions, allowing arbitrarily long and complex
sequences of lookups, restrictions and exceptions.  Postfix tries to
protect the administrator in as far as is reasonable, e.g.\ the restriction
\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
because the parameter it checks (the hostname given in the HELO command) is
under the control of the remote client.  Despite this, it is possible for
the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
all users --- the administrator must remember that \textit{With great power
comes great responsibility}.

Postfix uses simple lookup tables as the deciding factor when evaluating
some restrictions, e.g.\ in the example restriction below\newline
\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\newline The line above can be broken down as follows:

@@

The restriction \texttt{check\_client\_access} checks whether the \IP{}
address of the connected client is found in the file and returned the
associated action if found; the method of searching the file is dependant
on the type of the file (\texttt{cidr} in the example) --- see
\cite{postfix-lookup-tables} for more details.  Other restrictions
determine their result by consulting external sources, e.g.\
\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
address of the client is present in the \RBL{} \texttt{rbl.example.com},
rejecting the command if the client is listed.
This description is necessarily brief, for further details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.

@@

A \textit{policy server\/}~\cite{policy-servers} is an external program
consulted by Postfix to determine the fate of an \SMTP{} command.  The
policy server is given state information\footnote{Sample state information
is shown in table~\refwithpage{Example attributes sent to policy servers}}
and returns a verdict as described in \sectionref{Mixing and matching
Postfix restrictions}.  The policy server can perform more complex checks
than those provided by Postfix: a trivial example is restricting mail from
addresses associated with the payroll system to sending mail on the third
Tuesday after pay day only, to help prevent problems from spam or (worse)
phishing mails with faked sender addresses.\footnote{A
phishing mail might claim that the payroll system had a disastrous disk
failure; until the server is replaced and restored all salary payments will
have to be processed manually, so please reply to this mail with your name,
address and back account details.}

Some widely deployed policy servers:

@@

    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
        \SPF{}\label{spf introduction} records specify which mail servers
        are allowed to send mail claiming to be from a particular domain.
        The intention is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter} and
        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of \SMTP{}, e.g.\ forwarding mail from one company or
        university to another when a user changes jobs.

@@

    \item Using a scoring system such as
        Policyd-weight~\cite{policyd-weight} where tests accumulate points
        against the sending system --- if the eventual score is higher than
        a threshold the mail is rejected.  Postfix's restrictions are
        binary, and the administrator must manually whitelist clients if
        they are to bypass a restriction; using a threshold which requires
        hitting several restrictions frees the administrator from
        whitelisting clients which fall foul of one restriction only.

@@

\begin{table}[ht]

    \caption{Example attributes sent to policy servers}\label{Example
    attributes sent to policy servers}

@@

    \end{tabular}
\end{table}

@@

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that log files produced by Postfix are not fixed; they vary
widely from host to host, depending on the set of restrictions chosen by
the administrator.  With this in mind, one of the parser's design aims was
to make adding new rules as easy as possible, to enable administrators to
properly parse their log files.  To enable this the parser is divided into
three parts:

@@

    \item [rules] Rules match individual log lines and determine which
        actions will be executed.  Rules provide an easily extensible
        method of associating log lines with actions, and are described in
        detail in \sectionref{rules}.

    \item [actions] Actions are invoked to deal with a log line once it has
        been identified by the rules: actions modify data structures,
        handle complications, and cause data to be saved to the database.
        The actions perform the work of reconstructing the journey a mail
        takes through Postfix.  Full details of actions can be found in
        \sectionref{actions-in-detail} and \sectionref{adding new actions}.

    \item [framework] The framework is responsible for loading rules,
        managing data structures, finding the rule which matches each log
        line, invoking the correct action, etc.  The framework provides the
        structure which actions and rules plug into.  The framework is
        described in detail in \sectionref{framework}.  XXX EXTEND
        FRAMEWORK SECTION IF NECESSARY\@.

@@

XXX MERGE THE NEXT TWO PARAGRAPHS\@.

Decoupling the parsing rules from the associated actions and framework
allows new rules to be written and tested without requiring modifications
to the parser source code (significantly lowering the barrier to entry for
new or casual users who need to parse new log lines), and greatly
simplifies framework, actions and rules.  Decoupling also creates a clear
separation of functionality: rules handle low level details of identifying
log lines and extracting data from a log line; actions handle
the higher level details of following the path a mail takes through
Postfix, assembling the required data before storing it, dealing with
complications arising, etc; the framework provides services to actions and
stores data.

@@

XXX CLARIFY THE SENTENCE STARTING WITH "Although this may seem" IN THE NEXT
PARAGRAPH\@.

@@

\subsection{Conventions used in the document}

The words \textit{connection\/} and \textit{mail\/} are often used
interchangeably in this document; in general the word used was chosen based
on the context it appears in.

@@

Ten other parsers have been reviewed in
appendix~\refwithpage{other-parsers} as part of the background research for
this project.  None of the reviewed parsers perform the type of advanced
parsing and log correlation described here; all are intended to perform a
specific parsing and reporting task, rather than be a generic parser,
extracting data and leaving generation of reports from the data to other
programs.  Some parsers save the data extracted to a data store, but the
many discard all data once they have finished running, making historical
analysis impossible.  The other parsers reviewed all produce a report of
varying complexity and detail, whereas the program described here doesn't
attempt to produce a report at all; that responsibility is deferred to a
separate program, to be developed later.  The parsing algorithm and program
described here are designed to enable much more detailed log analysis by
providing a stable platform for subsequent programs to develop upon.

XXX EXTEND THIS\@.

@@

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
different from this parser: it attempts to present correlated log data in a
form suitable for a systems administrator to search using the myriad of
standard Unix text processing utilities already available, producing both a
\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
support both \CSV{} and Berkeley DB appears to have been a limiting factor:
\CSV{} is a very simple format where a record is stored in a single line,
with fields separated by a comma or other punctuation symbol.  Problems
with \CSV{} files include the difficulty in escaping separators, providing
multiple values for a field (e.g.\ multiple recipients) and adding new
fields.  There is no facility to include documentation of the fields,
separator, etc., unlike \SQL{} databases where every database includes a
schema naming the fields and the type of data they store (integer, text,
timestamp, etc.).

Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
unlike \SQL{} databases which store arbitrary tuples; in \LMA{}'s main
table the key is an integer referred to by secondary tables, and the value
used is a \CSV{} line containing all of the data for that row.  The
secondary by-sender, by-recipient, by-date and by-\IP{} tables use the
sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
referring to the main table is the value.  In effect this re-implements
\SQL{} foreign keys, but without the functionality offered by even the most
basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
requires custom code to search on some combination of the above, though the
authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
of the three output formats: it doesn't provide the functionality of a
basic \SQL{} database, not can it be used with standard Unix text
processing tools.

The data stored is limited to time and date, hostname and \IP{} address of
client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
codes or remote servers\footnote{A single mail may be sent to multiple
remote servers if it was addressed to recipients in different domains, or
Postfix needs to try multiple servers for one or more recipients.} is not
explained in the paper; experimental observation shows that multiple
records are added when there are multiple recipients (sadly the records are
not associated or linked in any way), and presumably the same approach is
taken when there are multiple destination servers.

The schema used with the MySQL database is undocumented, but at least it's
possible to discover the schema with an \SQL{} database, unlike with
Berkeley DB\@; all \SQL{} databases embed the schema into the database and
provide commands for displaying it.  Berkeley DB does not embed a schema,
as there is neither requirement nor benefit; it only provides \textbf{(key,
value)} pairs, so any additional structuring of the data is imposed by the
application, thus it behoves the application to document this structure.

@@

This section has provided background information on several topics relevant
to the remainder of the document.  It started with the motivation behind
the project, continuing with explanations of:

@@

    \item A description of the parser's novel design.

@@

    \item An in-depth comparison of this parser with \LMA{}, the parser
        described in previously published research.
